Solution of an Acoustic Transmission Inverse Problem by Extended Inversion

### 2020-12-07 repro/papers/discrep/paper.tex
current

\maketitle
\begin{abstract}
  Study of an extremely simple single-trace transmission example shows
  how an extended source formulation of full waveform inversion can
  produce an optimization problem without spurious local minima
  (``cycle skipping''), hence efficiently solvable via Newton-like
  local optimization methods. The data consist of a single trace
  recorded at a given distance from a point source. The velocity or
  slowness is presumed homogeneous, and the target source wavelet is
  presumed quasi-impulsive or localized near zero time lag. The source
  is extended by permitting energy to spread in time, and the spread
  is controlled by adding a weighted mean square of the extended
  source wavelet to the data misfit, to produce the extended inversion
  objective. The objective function and its gradient can be computed
  explicitly, and it is easily seen that the error in any local
  minimizer is bounded by a multiple of the permitted energy spread,
  which may be interpreted as a wavelength. The derivation shows several
  important features of all similar extended source algorithms. For
  example, nested optimization, with the source estimation in the
  inner optimization (variable projection method), is essential. The
  choice of the penalty operator, controlling the extended source
  degrees of freedom, is critical: in order to produce an objective
  immune from cycle-skipping, the penalty operator must be
  (pseudo-)differential. The penalty weight is chosen dynamically,
  according to a version of the Discrepancy Principle, in order to
  accelerate convergence to a near-global minimum. Like many other
  extended formulations, the example problem considered here permits
  the discrepancy criterion to be satisfied with zero penalty
  weight. This unusual feature removes a common obstacle to the use of
  the discrepancy principle, namely the determination of a
  satisfactory initial estimate of the penalty weight: the initial weight may
  be set to zero, and the dynamic algorithm updates the weight until
  convergence is achieved.

@@

\subsection{Proof of Theorem \ref{thm:fwi}}

%A natural formulation of the Problem Statement \ref{eqn:probstat0} is
%via least squares: that is, given $d \in D$, $\lambda >0$, find
%$(m,w) \in M \times \lW$ to minimize
%$\lerr: M \times W \rightarrow \bR$, defined in display
%\ref{eqn:redms}.  A pair $(m,w) \in M\times \lW$ satisfies the
%condition \ref{eqn:probstat0} iff $\lerr[m,w;d] \le \epsilon^2/2$, so a
%solution exists iff a global minimizer of $\lerr$ satisfies this
%inequality.

%In this discussion, I will presume that $d$ is 
%noise-free, as defined at the end of the last section (equation
%\ref{eqn:defdatanonoise}). Thus {\em a priori} a global minimizer of
%$e$ is a solution of the Problem Statement for any
%$\epsilon > 0$. The only
%question is how to find such a global minimizer. As mentioned in the
%introduction, the simple model problem studied here is a proxy for
%large-scale data fitting problems for which the only computationally
%feastible approach to minimization of analogs of $e$
%is some form of descent-based iterative local optimization. Any such
%algorithm requires the specification of an initial iterate
%$(m_0,w_0)$. Therefore the key question is: what
%constraints must be imposed on $(m_0,w_0)$ in order that the resulting
%sequence of iterates convertes to a global minimizer?

%\begin{theorem}
%  \label{thm:fwi}
%  Suppose that $0 <\lambda\le \lambda_{\rm max}$,  $m_* \in M, w_*
%  \in \lW, d=\lF[m_*]w_*$ is noise-free data per definition \ref{eqn:defdatanonoise},
%  Under assumption \ref{eqn:fullrec}, for any $m \in M$ with $r|m-m_*|>2\lambda$,
%\begin{equation}
%  \label{eqn:isovpm}
% \min_w e[m,w;d]=e[m,0,d]=\frac{1}{32\pi^2 r^2}\|w_*\|^2,
%\end{equation}
%and any such $(m,0)$ is a stationary point of $e$.
%\end{theorem}

%\begin{theorem}
%\label{thm:red}
%$\Ja[m,\cdot;d]$ is invertible:
%\begin{equation}
%\label{eqn:norminv}
%F[m]^TF[m] +\alpha^2 A^TA \in {\cal I}(W,W).
%\end{equation}
%(The superscript $T$ denotes the adjoint with respect to the inner products 
%in $W=L^2(\bR)$ and $D=L^2([t_{\rm min},t_{\rm max}])$.)
%Then the unique stationary point $\aw[m;d] \in W$ of
%$\Ja[m,\cdot;d]$ is a global minimizer, and is the solution of the {\em normal equation}:
%\begin{equation}
 % \label{eqn:norm}
  %(F[m]^TF[m]+\alpha^2A^TA)w= F[m]^Td, 
%\end{equation}
%The reduced or variable projection objective $\tJa$ defined in
%equation \ref{eqn:redexp} is also given by
%\begin{equation}
 % \label{eqn:redexp1}
 % \tJa[m;d] =\Ja[m,w_{\alpha}[m;d];d] = \min_w \Ja[m,w;d]
%\end{equation}
%Then $(m,w)$ is a local minimizer of $\Ja[\cdot,\cdot;d]$ if and only
%if $m$ is a local minimizer of $\tJa[\cdot;d]$ and $w=\aw[m;d]$.
%\end{theorem}

%\noindent{\bf Remark:} The choice
%of penalty weight $\alpha$ has a significant influence on the
%performance of numerical optimization algorithms in minimizing
%$\Ja$ and/or $\tJa$, but relatively little influence on the error estimates to
%follow. For the discussion of these estimates, we assume only that
%$\alpha>0$. Appendix A contains a description of an algorithm for
%constructive determination of $\alpha$.

\subsection{Proof of Theorem \ref{thm:norminv}}

%\begin{proof} of Theorem \ref{thm:norminv}: Since the normal operator
%  is invertible (item 1 of Proposition \ref{thm:norminvexp}), the
%  conclusions of this theorem follow from the Lax-Milgram Theorem
%  (\cite{Yosida}, III.7).
%\end{proof}

\subsection{Proof of Theorem \ref{thm:diffobj}}

\subsection{Proof of Theorem \ref{thm:boxcarbad}}

\subsection{Proof of Theorem \ref{thm:mnoiseres}}
\begin{proposition}
  \label{thm:rampgood}
  Suppose that
  \begin{enumerate}
  \item $m_* \in M$;
  \item $0 < \mu \le \lambda$, and $w_* \in W_{\mu}$;
  \item $d_* = F[m_*]w_*$;
  \item $a(t)=\min\{|t|,\tau\}$ in the definition \ref{eqn:annmult},
    with $\tau$ given by equation \ref{eqn:taudef}; and
  \item $\alpha > 0$.
  \end{enumerate}
  Then for any $m \in M$, 
  \begin{equation}
    | (m - m_*)r| > \lambda  \Rightarrow  \left|\frac{d}{dm}\tJa[m;d_*]\right| > \alpha^2 
    \frac{\lambda-\mu}{(1+(4\pi r)^2\alpha^2 
      (\lambda+\mu)^2)^{2}} \|w_*\|^2 
    \label{eqn:gradbndnonoise}
  \end{equation}
\end{proposition}
\begin{proof}
  As observed before, $\mbox{supp }\aw[m;d_*] \subset [t_{\rm
    min}-mr,t_{\rm max}-mr]\subset [-\tau,\tau]$, with $\tau$ defined
  in \ref{eqn:taudef}. Therefore, $a(t) = |t|$, $a a'(t) = t$ in the
  support of the integrand on the RHS of equation
  \ref{eqn:dexpjnonoise}, which therefore 
  becomes (after change of integration variable)
  %%%%%%%%%%%%%%%%%%%%%%%%%
  \begin{equation}
    \label{eqn:gradfinal}
    \frac{d}{dm}\tJa[m;d_*] = -r \alpha^2 \int \,dt \, 
  t(1+(4\pi r)^2 \alpha^2 
  t^2)^{-2}w_*(t+(m-m_*))^2.
  \end{equation}
  Recall that $w_*(t+(m-m_*)r)$
  vanishes if $|t+(m-m_*)r| > \lambda$. Therefore the integral on the
  RHS of equation \ref{eqn:gradfinal} can be re-written
  \[
    = -r\alpha^2\int_{-(m-m_*)r-\lambda}^{-(m-m_*)r+\lambda}
    \,dt\, t(1+(4\pi r)^2\alpha^2 t^2)^{-2}w_*\left(t+(m-m_*)r\right)^2
  \]
  Suppose that $\mu \le \lambda$ and $w_* \in W_{\mu}$. 
  If $m > m_*+\lambda/r$, then $t+(m-m_*)r \in \mbox{supp }w_*$
  implies $-\mu - \lambda < t < \mu-\lambda<0$, so 
  \[
    t(1+(4\pi r)^2\alpha^2 t^2)^{-2} < (\mu-\lambda)(1+(4\pi r)^2\alpha^2 (\mu+\lambda)^2)^{-2}<0
  \]
  in the support of the integrand in equation
  \ref{eqn:gradfinal}. Arguing similarly for $m<m_*-\lambda/r$, obtain
  a similar inequality, implying the conclusion \ref{eqn:gradbndnonoise}.
\end{proof}

@@

    % Recall that $\lE$ is the projection of $W$ onto $\lW$
%(i.e. multiplication by ${\bf 1}_{[-\lambda,\lambda]}$.

\subsection{Proof of Theorem \ref{thm:ipnoisesuf}}
\begin{proof} of Theorem \ref{thm:ipnoisesuf}:
From Theorem \ref{thm:norminv}, 
\[
  \aw[m;d](t) = (1+ (4 \pi r \alpha t)^2)^{-1}(w_*(t+(m-m_*)) + 4\pi r 
  n(t+mr)) 
\]
\[
  = \aw[m;d_*] + (1+ (4 \pi r \alpha t)^2)^{-1}4\pi r n(t+mr)
\]

>>>>

\maketitle
\begin{abstract}
  Study of an extremely simple single-trace transmission example shows
  how an extended source formulation of full waveform inversion can
  produce an optimization problem without spurious local minima
  (``cycle skipping''), hence efficiently solvable via Newton-like
  local optimization methods. The data consist of a single trace
  extracted from a causal pressure field, propagating in a fluid
  according to linear acoustics and recorded at a given distance from
  a transient point energy source. The velocity or slowness of the
  fluid is presumed homogeneous, and the target source intensity
  (``wavelet'') is presumed quasi-impulsive, with zero energy for time
  lags greater than a specified maximum lag (specified support). The
  data are assumed noisy, with known RMS signal-to-noise ratio. The
  inverse problem is: from the recorded trace, recover both the sound
  velocity and source wavelet with specified support, so that the data
  is fit with prescribed RMS relative error. After showing explicitly that
  the obvious least squares objective has multiple large residual
  minimizers, we introduce an extended approach. The source is extended by
  permitting energy to spread in time. The spread is controlled by
  adding a weighted mean square of the extended source wavelet to the
  data misfit, to produce the extended inversion objective. We choose
  a weight operator to penalize source energy away from zero lag via
  multiplication by the time lag. The resulting objective function and
  its gradient can be computed explicitly. We show that the error in 
  slowness component of any local minimizer is bounded by multiple of
  the maximum lag and the assumed noise level. In particular, there
  are no spurious local minima. We also show that the projection of
  the extended wavelet onto the interval of support yields a solution
  of the inverse problem as posed above, with appropriate choice of
  maxiumum lag and RMS relative error. In several appendices, we
  describe a constructive algorithm for controlling the penalty
  weight, the relation between maxiumum lag and RMS central frequency
  via the Heisenberg principle, and some structural details which the
  very simple inverse problem discussed here shares with more
  prototypical problems in industrial and academic seismology.

  %which may be interpreted as a wavelength.
  %The derivation shows
  %several important features of all similar extended source
  %algorithms. For example, nested optimization, with the source
  %estimation in the inner optimization (variable projection method),
  %is essential. The choice of the penalty operator, controlling the
  %extended source degrees of freedom, is critical: in order to produce
  %an objective immune from cycle-skipping, the penalty operator must
  %be (pseudo-)differential. The penalty weight is chosen dynamically,
  %according to a version of the Discrepancy Principle, in order to
  %accelerate convergence to a near-global minimum. Like many other
  %extended formulations, the example problem considered here permits
  %the discrepancy criterion to be satisfied with zero penalty
 % weight. This unusual feature removes a common obstacle to the use of
  %the discrepancy principle, namely the determination of a
  %satisfactory initial estimate of the penalty weight: the initial
 % weight may be set to zero, and the dynamic algorithm updates the
 % weight until convergence is achieved.

@@

\begin{proposition}
  \label{thm:rampgood}
  Suppose that
  \begin{enumerate}
  \item $m_* \in M$;
  \item $0 < \mu \le \lambda$, and $w_* \in W_{\mu}$;
  \item $d_* = F[m_*]w_*$;
  \item $a(t)=\min\{|t|,\tau\}$ in the definition \ref{eqn:annmult},
    with $\tau$ given by equation \ref{eqn:taudef}; and
  \item $\alpha > 0$.
  \end{enumerate}
  Then for any $m \in M$, 
  \begin{equation}
    | (m - m_*)r| > \lambda  \Rightarrow  \left|\frac{d}{dm}\tJa[m;d_*]\right| > \alpha^2 
    \frac{\lambda-\mu}{(1+(4\pi r)^2\alpha^2 
      (\lambda+\mu)^2)^{2}} \|w_*\|^2 
    \label{eqn:gradbndnonoise}
  \end{equation}
\end{proposition}
\begin{proof}
  As observed before, $\mbox{supp }\aw[m;d_*] \subset [t_{\rm
    min}-mr,t_{\rm max}-mr]\subset [-\tau,\tau]$, with $\tau$ defined
  in \ref{eqn:taudef}. Therefore, $a(t) = |t|$, $a a'(t) = t$ in the
  support of the integrand on the RHS of equation
  \ref{eqn:dexpjnonoise}, which therefore 
  becomes (after change of integration variable)
  %%%%%%%%%%%%%%%%%%%%%%%%%
  \begin{equation}
    \label{eqn:gradfinal}
    \frac{d}{dm}\tJa[m;d_*] = -r \alpha^2 \int \,dt \, 
  t(1+(4\pi r)^2 \alpha^2 
  t^2)^{-2}w_*(t+(m-m_*))^2.
  \end{equation}
  Recall that $w_*(t+(m-m_*)r)$
  vanishes if $|t+(m-m_*)r| > \lambda$. Therefore the integral on the
  RHS of equation \ref{eqn:gradfinal} can be re-written
  \[
    = -r\alpha^2\int_{-(m-m_*)r-\lambda}^{-(m-m_*)r+\lambda}
    \,dt\, t(1+(4\pi r)^2\alpha^2 t^2)^{-2}w_*\left(t+(m-m_*)r\right)^2
  \]
  Suppose that $\mu \le \lambda$ and $w_* \in W_{\mu}$. 
  If $m > m_*+\lambda/r$, then $t+(m-m_*)r \in \mbox{supp }w_*$
  implies $-\mu - \lambda < t < \mu-\lambda<0$, so 
  \[
    t(1+(4\pi r)^2\alpha^2 t^2)^{-2} < (\mu-\lambda)(1+(4\pi r)^2\alpha^2 (\mu+\lambda)^2)^{-2}<0
  \]
  in the support of the integrand in equation
  \ref{eqn:gradfinal}. Arguing similarly for $m<m_*-\lambda/r$, obtain
  a similar inequality, implying the conclusion \ref{eqn:gradbndnonoise}.
\end{proof}

@@

\begin{proof} of Theorem \ref{thm:ipnoisesuf}:
From Theorem \ref{thm:norminv}, 
\[
  \aw[m;d](t) = (1+ (4 \pi r \alpha t)^2)^{-1}(w_*(t+(m-m_*)) + 4\pi r 
  n(t+mr)) 
\]
\[
  = \aw[m;d_*] + (1+ (4 \pi r \alpha t)^2)^{-1}4\pi r n(t+mr)
\]
